---
title: "Survey Analysis with R - Inferential Statistics"
author: "Jemma Nelson and Rian DeFaccio"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

# Survey Statistics Tests

This section introduces the idea of hypothesis testing and then covers
common inferential statistical tests used in survey analysis with
complex survey designs. These tests account for survey weights and
design effects, providing valid inference for population estimates.

## Libraries

```{r}
library(tidyverse)
library(survey)
library(srvyr)
```

## Data

Here we setup a design object for the Residential Energy Consumption
Survey data which we first saw in the prior lesson.

```{r}
recs_des <- read.csv('../data/recs_2020.csv') %>%
  as_survey_rep(
    weights = NWEIGHT,
    repweights = NWEIGHT1:NWEIGHT60,
    type = "JK1",
    scale = 59 / 60,
    mse = TRUE
  )
```

# Statistical Inference and Hypothesis Testing

## Introduction to Statistical Inference

------------------------------------------------------------------------

Statistical inference is the process of using sample data to make
conclusions about a larger population. Instead of measuring every
individual in a population (which is often impossible), we collect a
representative sample and use statistical methods to draw reasonable
conclusions.

Key concepts include:

-   **Population**: The entire group we want to study
-   **Sample**: A subset of the population we actually observe
-   **Parameter**: A numerical characteristic of the population
    (unknown)
-   **Statistic**: A numerical characteristic of the sample (calculated
    from data)

## The Logic of Hypothesis Testing

------------------------------------------------------------------------

Hypothesis testing provides a structured framework for making decisions
about population parameters based on sample evidence. The process
involves:

**Setting up hypotheses:**

-   **Null hypothesis (H₀)**: The default assumption or "no effect"
    claim
-   **Alternative hypothesis (H₁ or Hₐ)**: The claim we're testing for

**The testing process:**

-   Assume the null hypothesis is true
-   Calculate how likely our observed data would be under this
    assumption
-   If the data would be very unlikely under H₀, we reject the null
    hypothesis
-   If the data is reasonably likely under H₀, we fail to reject the
    null hypothesis

## Understanding P-Values

------------------------------------------------------------------------

A **p-value** is the probability of observing data as extreme or more
extreme than what we actually observed, assuming the null hypothesis is
true.

**Key characteristics of p-values:**

-   Always between 0 and 1
-   Smaller p-values indicate stronger evidence against the null
    hypothesis
-   A p-value is NOT the probability that the null hypothesis is true
-   Common threshold: p \< 0.05 is often used to determine "statistical
    significance"

**Interpreting p-values:**

-   p = 0.03: "If the null hypothesis were true, we'd see results this
    extreme only 3% of the time"
-   p = 0.15: "If the null hypothesis were true, we'd see results this
    extreme 15% of the time"
-   p = 0.001: "If the null hypothesis were true, we'd see results this
    extreme only 0.1% of the time"

## Steps in Hypothesis Testing

------------------------------------------------------------------------

1.  **State the hypotheses** (H₀ and H₁)
2.  **Choose significance level** (α, typically 0.05)
3.  **Select appropriate test statistic** (t-test, z-test, chi-square,
    etc.)
4.  **Calculate the test statistic** from sample data
5.  **Find the p-value** or critical value
6.  **Make a decision** (reject or fail to reject H₀)
7.  **State conclusion** in context of the problem

## Common Misconceptions

------------------------------------------------------------------------

**About p-values:**

-   A p-value does NOT tell us the probability that our hypothesis is
    correct
-   p \< 0.05 doesn't mean the result is "important" or "large"
-   Statistical significance ≠ practical significance

**About hypothesis testing:**

-   We never "prove" a hypothesis true
-   We never "accept" the null hypothesis, only fail to reject it
-   Small sample sizes can lead to low statistical power

## Example Application

------------------------------------------------------------------------

**Research question:** Does a new teaching method improve test scores?

-   **H₀**: The new method has no effect (μ₁ = μ₂)
-   **H₁**: The new method improves scores (μ₁ \> μ₂)
-   **Data**: Sample of students using new method vs. traditional method
-   **Result**: p = 0.02
-   **Conclusion**: We reject H₀ and conclude the new method appears to
    improve test scores

## Key Takeaways

------------------------------------------------------------------------

Statistical inference allows us to make informed decisions about
populations based on sample data. Hypothesis testing provides a
systematic approach, but requires careful interpretation of results.

Remember:

-   P-values measure evidence against the null hypothesis
-   Statistical significance doesn't guarantee practical importance
-   Always consider the context and limitations of your analysis
-   Replication and effect size are crucial for drawing robust
    conclusions

# Overview of Survey Statistical Tests

Now let's begin to apply all of that theory! Here are the statistical
tests we will be exploring, and which functions we will use to implement
them.

## Tests for Continuous Variables

| Test Type | Function(s) | Purpose | Example Use Case |
|----------------|----------------|---------------------|---------------------|
| Comparison of Means | `survey::svyttest()` | Test if mean of a continuous variable equals a value (one-sample) or differs between groups (two-sample), accounting for survey design | Compare average electricity bills between households with and without air conditioning |
| Comparison of Proportions | `survey::svyttest()` | Test if a binary variable's proportion equals a value or differs between groups, accounting for survey design | Test if the proportion of households with air conditioning differs from 90% |

## Tests for Categorical Variables

| Test Type | Function(s) | Purpose | Example Use Case |
|----------------|----------------|---------------------|---------------------|
| Goodness-of-fit Test | `survey::svygofchisq()` | Assess if observed distribution of a categorical variable matches expected proportions | Test if housing unit type distribution matches Census data |
| Test of Independence | `survey::svychisq()` | Evaluate if two categorical variables are independent in the population, adjusting for survey design | Test if housing unit type is associated with space heating type |
| Test of Homogeneity | `survey::svychisq()` | Compare distribution of a categorical variable across groups, accounting for survey sampling | Test if housing unit type distribution is the same across climate regions |

These functions from the {survey} and {srvyr} packages ensure that
statistical inference properly reflects the survey's complex sampling
structure.

# Comparisons of Means

In survey analysis, comparing a continuous variable to a single value or
to another continuous variable is commonly done using survey-adjusted
t-tests. These tests account for the survey design, ensuring valid
inference.

-   **One-sample t-test:** Tests if the mean of a continuous variable
    differs from a specified value.
-   **Two-sample t-test:** Compares means between two groups.
-   **Paired t-test:** Compares two related measurements within the same
    units (e.g., before and after).

The following examples demonstrate how to perform these tests using the
`svyttest()` function from the {survey} or {srvyr} packages.

## One-Sample t-test: Testing Against a Known Value

A one-sample t-test is used to determine whether the mean of a single
continuous variable differs from a specified value. In survey analysis,
the `svyttest()` function adjusts for survey design and weights.

For example, to test whether the mean nighttime summer temperature
(`SummerTempNight`) differs from 68°F:

A one-sample t-test evaluates whether the population mean differs
significantly from a hypothesized value. It answers questions like "Is
the average summer temperature different from 68°F?"

With survey data, we use `svyttest()` with the formula
`variable - hypothesized_value ~ 0`.

```{r}
# One-sample t-test example
one_sample_test <- recs_des %>%
  svyttest(
    formula = SummerTempNight - 68 ~ 0, # Testing if mean equals 68°F
    design = .,
    na.rm = TRUE
  )

# Display test results
one_sample_test
```

The output shows the test statistic, degrees of freedom, and p-value. A
small p-value (\< 0.05) indicates that the mean temperature is
significantly different from 68°F. We can also extract the confidence
interval:

```{r}
# Extract confidence interval
confint(one_sample_test)
```

```{r}
# Extract confidence interval at different significance threshold
confint(one_sample_test, level = 0.80)
```

## Two-Sample t-test: Comparing Groups

A two-sample t-test compares means between two independent groups. It
answers questions like "Do households with air conditioning have
different electricity bills than those without?"

For survey data, we use `svyttest()` with the formula
`continuous_variable ~ grouping_variable`.

```{r}
# Two-sample t-test comparing electricity bills between households with and without AC
two_sample_test <- recs_des %>%
  svyttest(
    formula = DOLLAREL ~ ACUsed,  # Comparing DOLLAREL between ACUsed groups
    design = .,
    na.rm = TRUE
  )

# Display test results
two_sample_test
```

```{r}
# Visualize group means
recs_des %>%
  group_by(ACUsed) %>%
  summarize(mean_bill = survey_mean(DOLLAREL, vartype = "ci", na.rm = TRUE)) %>%
  knitr::kable(digits = 2)
```

In this example, the p-value tells us whether the difference in
electricity bills between households with and without air conditioning
is statistically significant. The confidence interval of the difference
helps us understand the magnitude of this difference.

## Paired Sample t-test: Comparing Measurements on the Same Units

A paired t-test compares two related measurements, like summer vs.
winter temperatures in the same households. We can test this using a
one-sample test on the differences:

```{r}
# Paired t-test comparing summer and winter temperatures
paired_test <- recs_des %>%
  svyttest(
    formula = SummerTempNight - WinterTempNight ~ 0,  # Testing if the difference equals 0
    design = .,
    na.rm = TRUE
  )

paired_test
```

# Testing Proportions

Proportions tests with survey data follow similar principles to means
tests but work with binary outcomes. The `svyttest()` function can be
used for proportion tests by converting categorical variables to numeric
indicators (0/1).

## One-Sample Proportion Test

A one-sample proportion test determines whether a population proportion
differs from a specified value, accounting for the survey design.

```{r}
# First, let's examine the proportion of households with air conditioning
acprop <- recs_des %>%
  group_by(ACUsed) %>%
  summarize(p = survey_prop())

# Display the results
acprop
```

To test if the proportion of households with air conditioning differs
from 90%:

```{r}
# One-sample proportion test
prop_test_one <- recs_des %>%
  svyttest(
    formula = (ACUsed == TRUE) - 0.90 ~ 0,  # Testing if proportion equals 90%
    design = .,
    na.rm = TRUE
  )

# Display raw results
prop_test_one
```

```{r}
# For cleaner output with formatted p-values
broom::tidy(prop_test_one) %>%
  mutate(
    p.value = ifelse(p.value < 0.001, "<0.001", p.value)
  ) %>%
  knitr::kable(
    digits = 3,
    format.args = list(big.mark = ",", nsmall = 3)
  )
```

In this test:

-   The formula `(ACUsed == TRUE) - 0.90 ~ 0` converts the logical
    variable to 0/1 and tests if it differs from 0.90
-   A significant p-value indicates that the true proportion differs
    from 90%
-   The estimate shows the difference between the observed proportion
    and 90%

## Two-Sample Proportion Test

To compare proportions between two groups, we can use a similar approach
to two-sample means tests.

```{r}
recs_des %>%
  group_by(ACUsed) %>%
  summarize(mean = survey_mean(DOLLAREL, na.rm = TRUE)) %>%
  knitr::kable()
```

```{r}
# Compare proportions of households with solar panels between regions
region_solar_test <- recs_des %>%
  svyttest(
    formula = DOLLAREL ~ ACUsed,  # Comparing solar panel adoption by region
    design = .,
    na.rm = TRUE
  )

# Display results
region_solar_test
```

# Categorical Variable Tests

This section demonstrates how to test hypotheses about categorical
variables in survey data, using survey-adjusted chi-square tests. These
include:

-   **Goodness-of-fit tests:** Do observed category frequencies match
    expected values?
-   **Tests of independence:** Are two categorical variables associated?
-   **Tests of homogeneity:** Is the distribution of a categorical
    variable the same across groups?

We use the `svygofchisq()` and `svychisq()` functions from the {survey}
package, which properly account for survey weights and design effects.

## Goodness-of-Fit Test

A goodness-of-fit test evaluates whether an observed categorical
distribution matches an expected distribution. This is useful for
comparing survey data to known population values.

When performing inferential statistical tests, it is important to handle
missing data appropriately. Most survey analysis functions in R,
including `svygofchisq()`, require complete data for the variables
involved in the test. If any observations have missing values for the
tested variable, they are excluded from the analysis. This is typically
done using `drop_na()` or by specifying `na.rm = TRUE` in the function
call.

Dropping missing data ensures that the test statistic is calculated only
on valid, observed responses. Including missing values would either
cause errors or bias the results, as the statistical methods cannot
infer the category for missing responses. Therefore, before running a
goodness-of-fit test (or any inferential test), we remove observations
with missing values for the relevant variable to ensure accurate and
interpretable results.

```{r}
# First, examine the distribution of educational attainment
recs_des %>%
  drop_na(HousingUnitType) %>%
  group_by(HousingUnitType) %>%
  summarize(p = survey_mean()) %>%
  mutate(p_se = ifelse(p_se < 0.001, "<0.001", p_se)) %>%
  knitr::kable(
    digits = 3,
    format.args = list(big.mark = ",", nsmall = 3)
  )

# Test if education distribution matches expected values (e.g., from Census data)
gof_test <- recs_des %>%
  svygofchisq(
    formula = ~ HousingUnitType,
    p = c(0.10, 0.18, 0.05, 0.06, 0.61),  # Expected proportions
    design = .,
    na.rm = TRUE
  )

# Display test results
gof_test
```

In this goodness-of-fit test:

-   The `formula` specifies the categorical variable to test
-   The `p` argument provides the expected proportions (which must sum
    to 1)
-   A significant p-value indicates that the observed distribution
    differs from the expected distribution
-   The test accounts for the survey design when calculating the
    chi-square statistic

## Test of Independence

A test of independence determines whether two categorical variables are
associated within the population, accounting for the survey design. For
example, we might want to know if the type of housing unit is related to
the type of space heating used.

The `svychisq()` function from the {survey} package performs a
survey-adjusted chi-square test of independence. The formula combines
the two categorical variables with a `+` sign.

A significant p-value from this test suggests that the variables are not
independent, indicating an association between them. To better
understand the relationship, we can also create a cross-tabulation
showing the proportion of each combination of categories.

```{r}
# do a crosstab first
housing_heat_crosstab <- recs_des %>%
    group_by(HousingUnitType, SpaceHeatingUsed) %>%
    summarize(prop = survey_mean()) %>%
    select(-prop_se) %>%
    tidyr::pivot_wider(
        names_from = HousingUnitType,
        values_from = prop
    ) %>%
    knitr::kable(digits = 3)

housing_heat_crosstab
```

```{r}
# Test independence between HousingUnitType and SpaceHeatingUsed
independence_test <- recs_des %>%
    svychisq(
        formula = ~ HousingUnitType + SpaceHeatingUsed,
        design = .,
        statistic = "Wald",
        na.rm = TRUE
    )

# Display test results
independence_test
```

In this independence test:

-   The `formula` combines two categorical variables with `+` to test
    their independence
-   The `statistic = "Wald"` option is appropriate for complex survey
    designs
-   A significant p-value suggests the variables are not independent
    (i.e., they are associated)
-   The cross-tabulation helps interpret the nature of the relationship

## Test of Homogeneity

A test of homogeneity evaluates whether the distribution of a
categorical variable is the same across different groups. While
technically similar to independence tests, these assess if a categorical
variable's distribution varies across population subgroups.

```{r}
# Create a cross-tabulation showing party distribution by region
tab <- recs_des %>%
  group_by(HousingUnitType, ClimateRegion_BA) %>%
  summarize(prop = survey_mean()) %>%
  select(-prop_se) %>%
  tidyr::pivot_wider(
        names_from = HousingUnitType,
        values_from = prop
    ) %>%
  knitr::kable(digits = 3)

tab
```

```{r}
table(recs_des$variables$ClimateRegion_BA)

# Test if political party affiliation distribution is the same across regions
homogeneity_test <- recs_des %>%
  svychisq(
    formula = ~ HousingUnitType + ClimateRegion_BA,  # Testing if HousingUnitType is the same across regions
    design = .,
    statistic = "Chisq",
    na.rm = TRUE
  )

# Display test results
homogeneity_test
```

In this homogeneity test:

-   The formula structure is the same as for independence tests
-   We're specifically asking if the distribution of party affiliation
    is consistent across regions
-   A significant p-value indicates the distribution differs by region
-   The proportion table allows us to see how party distribution varies
    across regions

These three chi-square-based tests form the foundation for analyzing
categorical data in complex surveys, allowing us to make
population-level inferences while accounting for the survey design.

## Homogeneity Test vs. Independence Test

While both the test of homogeneity and the test of independence use the
chi-square statistic and have similar formula syntax in R, they answer
different research questions:

|   | Test of Independence | Test of Homogeneity |
|------------------------|------------------------|------------------------|
| **Purpose** | Determines whether two categorical variables are associated within a single population. | Assesses whether the distribution of a categorical variable is the same across different populations or groups. |
| **Example** | Is housing unit type associated with space heating type in the overall survey population? | Is the distribution of housing unit types the same across different climate regions? |
| **Interpretation** | A significant result suggests a relationship (dependence) between the variables. | A significant result indicates that at least one group has a different distribution compared to others. |

In practice, both tests use similar code and statistical methods, but
the context and interpretation differ:

-   Independence tests focus on association within a single population.
-   Homogeneity tests compare distributions across multiple populations
    or groups.

# Conclusion

Survey data analysis requires specialized statistical methods to account
for complex sampling designs, weights, and clustering. The {survey} and
{srvyr} packages in R provide robust tools for conducting valid
inferential tests—such as t-tests and chi-square tests—on means,
proportions, and categorical variables. By using these survey-adjusted
methods, analysts can draw accurate, population-level conclusions from
survey data, ensuring that results reflect the true variability and
structure of the sampled population. Mastery of these techniques is
essential for anyone working with survey data sets in applied research
or policy analysis.

Some key concepts from this lesson include:

-   Always use survey-adjusted statistical tests (e.g., `svyttest()`,
    `svychisq()`) to account for weights and complex sampling.
-   Interpret results in the context of the survey design—standard
    errors and p-values reflect the population, not just the sample.
-   Use the {survey} and {srvyr} packages for streamlined, reproducible
    analysis workflows.
-   Carefully check variable coding and missing data before running
    inferential tests.
