---
title: "GitHub and Source Control in RStudio"
author: "Rian DeFaccio for BH Analytics, Copyright BH Analytics, 2026"
date: "`r Sys.Date()`"
format: pdf
---

## Overview

------------------------------------------------------------------------

In this lesson we will cover using Git and GitHub as part of a reproducible analysis workflow inside RStudio. Topics include:

-   Why version control matters for data analysis
-   Installing and configuring Git
-   Connecting RStudio to GitHub with a Personal Access Token (PAT)
-   Core Git concepts: repositories, commits, branches
-   Using the RStudio Git pane to stage, commit, push, and pull
-   Creating and cloning repositories
-   Managing `.gitignore`
-   Common pitfalls and how to recover from them

## Why Version Control?

------------------------------------------------------------------------

If you have ever saved files named `analysis_final.R`, `analysis_final_v2.R`, or `analysis_ACTUALLY_final.R`, you already understand the problem that version control solves.

**Git** is a version control system that tracks every change you make to a file over time. With Git you can:

-   See exactly what changed, when, and why (via commit messages).
-   Revert any file (or your entire project) back to any earlier state.
-   Work on experimental ideas in a **branch** without breaking working code.
-   **Collaborate** with colleagues without overwriting each other's work.

**GitHub** is a website that hosts Git repositories in the cloud. It acts as the remote backup of your project and the central hub for collaboration.

::: callout-tip
## 🔁 SAS vs. R: Version Control Culture

SAS shops historically relied on shared network drives and manual naming conventions (`prog_v1.sas`, `prog_final.sas`) to manage code history. Formal version control for analytical code was rare.

Git and GitHub are standard practice in the R and open-source data science world. RStudio (Posit) has built-in Git integration, and tools like `usethis` make setup nearly effortless. Adopting version control is one of the highest-leverage habits you can build as an R analyst.
:::

## Setup and Configuration

------------------------------------------------------------------------

### Step 1 — Install Required R Packages

The `usethis` and `gitcreds` packages are used throughout this lesson. Neither is part of base R, so install them before proceeding:

```{r eval=FALSE}
install.packages(c("usethis", "gitcreds"))
```

### Step 2 — Install Git

Git is a separate program from R and RStudio and must be installed first.

-   **Windows:** Download from 🌐 <https://git-scm.com/download/win> and run the installer with default settings.
-   **Mac:** Run `git --version` in the Terminal; macOS will prompt you to install Command Line Tools if Git is not present.
-   **Linux:** `sudo apt install git` (Debian/Ubuntu) or equivalent.

After installing, restart RStudio. You can verify RStudio found Git under **Tools → Global Options → Git/SVN**.

### Step 3 — Configure Your Identity

Git records your name and email with every commit. Run these lines **once** in the R console (or a terminal):

```{r eval=FALSE}
# Replace with your own name and email
usethis::use_git_config(
  user.name  = "Rian DeFaccio",
  user.email = "rdefaccio@gmail.com"
)
```

### Step 4 — Connect RStudio to GitHub (Personal Access Token)

GitHub no longer accepts passwords over HTTPS. Instead, you authenticate with a **Personal Access Token (PAT)**. The `usethis` package makes this easy:

```{r eval=FALSE}
library(usethis)

# Step 1: Opens GitHub in your browser to create a token.
#         Choose the "repo", "workflow", and "gist" scopes at minimum.
create_github_token()

# Step 2: Opens a secure prompt where you paste the token.
#         gitcreds saves it to your OS credential store.
gitcreds::gitcreds_set()
```

::: callout-note
## Where is the PAT actually stored?

`gitcreds::gitcreds_set()` saves your token to your **operating system's secure credential store** — Windows Credential Manager, macOS Keychain, or the Linux Secret Service. The token lives at the OS level, completely outside your R project folder, and is never written to any file you could accidentally commit.

This is why you should **not** store a PAT in:

-   An R script (`.R`) or Quarto document (`.qmd`) — these can be committed to Git
-   Your `.Renviron` file — this is plain text on disk and easy to expose accidentally

Once stored via `gitcreds`, RStudio and Git discover it automatically in the background. You never need to reference it in code.
:::

Once set, RStudio can push and pull to GitHub without prompting you for credentials on every action.

## Core Git Concepts

------------------------------------------------------------------------

| Concept | What it is |
|------------------------------------|------------------------------------|
| **Repository (repo)** | A folder tracked by Git. Contains your files plus a hidden `.git/` folder that stores the full history. |
| **Commit** | A saved snapshot of your staged changes, paired with a message describing what changed and why. |
| **Branch** | An independent line of development. The default branch is usually called `main`. |
| **Remote** | A copy of your repo hosted elsewhere — typically GitHub (`origin`). |
| **Push / Pull** | `push` sends your local commits to the remote; `pull` downloads changes from the remote into your local repo. |

### The Basic Git Workflow

Every day-to-day interaction follows the same simple loop:

1.  **Edit** — make changes to your files in RStudio.
2.  **Stage** — check the box next to files you want to include in the next commit.
3.  **Commit** — write a short message and save the snapshot locally.
4.  **Push** — upload your commits to GitHub.

## Creating and Cloning Repositories

------------------------------------------------------------------------

### Option A — Start on GitHub, clone into RStudio

This is the recommended starting point for new projects.

1.  Go to 🌐 <https://github.com> and click **New repository**.
2.  Give it a name, choose Public or Private, and tick **Add a README file**.
3.  Click the green **Code** button and copy the HTTPS URL.
4.  In RStudio: **File → New Project → Version Control → Git**, paste the URL, and choose a local folder.

RStudio will clone the repository and open it as a new project with Git integration already active.

### Option B — Start locally, push to GitHub

If you already have an R project, `usethis` can initialize Git and create the GitHub repo in one step:

NOTE — need to have an R Project initialized...

```{r eval=FALSE}
library(usethis)

# Initialize a local Git repo (if not done yet)
use_git()

# Create a GitHub repo and push (requires PAT setup from Step 3 above)
use_github()
```

## The RStudio Git Pane

------------------------------------------------------------------------

Once a project is version-controlled, RStudio shows a **Git** tab in the upper-right pane (next to Environment and History). This is your everyday interface.

### Key Elements

| Element | Description |
|------------------------------------|------------------------------------|
| **Staged checkbox** | Tick to include a file in the next commit (equivalent to `git add`). |
| **Status icons** | `M` = Modified, `A` = Added (new), `D` = Deleted, `?` = Untracked |
| **Commit button** | Opens the commit dialog where you write your message and finalize the snapshot. |
| **Push / Pull** | Blue arrow (↓) pulls from GitHub; green arrow (↑) pushes to GitHub. |
| **Diff viewer** | Click a file name to see exactly what lines changed (green = added, red = removed). |
| **History** | Clock icon — shows every past commit, who made it, when, and what changed. |

### Writing Good Commit Messages

A commit message should complete the sentence *"When applied, this commit will…"*

```         
# Good
Add age-adjusted mortality rates to Table 2

# Too vague
Update stuff

# Too long
Changed the mortality rate calculation in the analysis script to use
age-adjusted rates instead of crude rates because the reviewer asked
for that in their comments from last week's meeting
```

A short imperative phrase (50 characters or fewer) is the convention.

## Branching and Pull Requests *(Optional)*

------------------------------------------------------------------------

For solo work, committing directly to `main` is fine. For team projects — or to safely develop a new feature — use a **branch**.

```{r eval=FALSE}
# Create and switch to a new branch from the R console
usethis::pr_init("add-smoking-analysis")
```

You can also create a branch from the RStudio Git pane by clicking the branch icon (looks like a diamond with arrows).

When your work on the branch is ready, open a **Pull Request (PR)** on GitHub. A PR is a proposal to merge your branch back into `main`. It gives teammates a chance to review your changes before they become part of the official repo.

```{r eval=FALSE}
# Push your branch and open a draft PR on GitHub
usethis::pr_push()
```

## Managing `.gitignore`

------------------------------------------------------------------------

Not everything in your project folder should be tracked. The `.gitignore` file tells Git which files and folders to ignore completely.

Common things to exclude in an R project:

```         
# R artifacts
.Rhistory
.RData
.Rproj.user/

# Rendered outputs (regenerate from source instead)
*.html
*.pdf

# Large or sensitive data files
data/
*.csv
*.xlsx

# Credentials — NEVER commit these
.env
secrets.R
```

::: callout-caution
## Keep credentials out of Git

If you accidentally commit a file containing a password, API key, or PAT, treat it as compromised immediately — rotate the credential even after you delete the file from history. Git history is permanent unless forcibly rewritten, and public repos are indexed by bots.
:::

RStudio creates a sensible `.gitignore` automatically for new R projects. You can also add patterns using:

```{r eval=FALSE}
usethis::use_git_ignore(c("data/", "*.csv"))
```

## Common Pitfalls

------------------------------------------------------------------------

### Merge Conflicts

A **merge conflict** happens when two people (or two branches) edit the same lines of the same file and Git cannot automatically decide which version to keep.

Git marks the conflict directly inside the file:

```         
# <<<<<<< HEAD
result <- model_v1 |> predict(new_data)
=======
result <- model_v2 |> predict(new_data)
# >>>>>>> feature-branch
```

To resolve: edit the file to keep whichever version (or combination) is correct, delete the conflict markers, stage the file, and commit.

The best prevention is **pull before you push** — always pull the latest changes from GitHub before starting a work session.

### Accidentally Committed a Large File

If you commit a large data file before adding it to `.gitignore`, use:

```{r eval=FALSE}
# Remove a file from tracking without deleting it from disk
system("git rm --cached data/large_file.csv")
```

Then add the path to `.gitignore` and commit the change.

### Lost Commits / Detached HEAD

If you end up in a "detached HEAD" state or feel like you've lost work, do not panic and do not force-push. Run:

```{r eval=FALSE}
system("git reflog")
```

`reflog` is Git's internal safety net — it records every position `HEAD` has been at, including commits that appear lost. You can recover from nearly any mistake as long as you haven't run `git gc`.

## Hands-On Exercise

------------------------------------------------------------------------

Practice the complete workflow end-to-end:

1.  **Create a repo** — Create a new public repository on GitHub named `r-version-control-practice` with a README.

2.  **Clone it** — Clone the repo into a new RStudio project using **File → New Project → Version Control → Git**.

3.  **Add a script** — Create a new R script (`practice.R`) that loads `ggplot2` and makes a simple plot.

4.  **Stage and commit** — In the Git pane, tick the checkbox next to `practice.R`, click Commit, write a message like `"Add initial practice script"`, and click Commit.

5.  **Push** — Click the Push button (↑) and confirm the commit appears on GitHub in your browser.

6.  **Edit and push again** — Modify the plot (change the color, add a title), stage, commit with a new message, and push.

7.  **View history** — Click the History (clock) icon in the Git pane and confirm you can see both commits.

## Quick Reference

------------------------------------------------------------------------

| Action | RStudio UI | `usethis` / terminal equivalent |
|------------------------|------------------------|------------------------|
| Initialize Git in existing project | Tools → Project Options → Git/SVN | `usethis::use_git()` |
| Clone a repo | New Project → Version Control | `git clone <url>` |
| Stage a file | Tick checkbox in Git pane | `git add <file>` |
| Commit | Commit button → write message | `git commit -m "message"` |
| Push | ↑ arrow in Git pane | `git push` |
| Pull | ↓ arrow in Git pane | `git pull` |
| Create branch | Branch icon in Git pane | `usethis::pr_init("name")` |
| Check status | Git pane file list | `git status` |
| View history | Clock icon in Git pane | `git log` |

## Further Resources

------------------------------------------------------------------------

-   🌐 Happy Git and GitHub for the useR (the definitive R-focused guide): <https://happygitwithr.com/>
-   🌐 `usethis` package documentation: <https://usethis.r-lib.org/>
-   🌐 GitHub Docs — Getting started: <https://docs.github.com/en/get-started>
-   🌐 Pro Git book (free, comprehensive): <https://git-scm.com/book/en/v2>
