---
title: "Survey Analysis with R - Descriptive Statistics Lab SOLUTIONS"
author: "Instructor Solutions"
format: pdf
---

## Survey Descriptive Statistics Lab

In this lab, you will practice computing descriptive statistics for survey data using the ANES 2020 dataset. You will learn to calculate survey-weighted estimates including counts, sums, means, proportions, and quantiles while accounting for the complex survey design.

Please refer to the codebook when looking up information on variables:

> 🌐[ANES 2020 Time Series Study
Full Release](https://electionstudies.org/wp-content/uploads/2022/02/anes_timeseries_2020_userguidecodebook_20220210.pdf)

### 1. Load Libraries and Data

Load the required packages and import the ANES 2020 dataset.

```{r}
# LOAD REQUIRED LIBRARIES
library(tidyverse)
library(survey)
library(srvyr)

# remove scientific notation
options(scipen = 999)
```

```{r}
# IMPORT THE ANES 2020 DATa
# SAVE AS: anes_data
anes_data <- read.csv('../data/anes_2020.csv')
```

```{r}
# EXAMINE THE BASIC STRUCTURE OF THE DATA
# DISPLAY: dimensions and first few rows
dim(anes_data)
head(anes_data, 3)
```

### 2. Create Survey Design Object

```{r}
# CREATE A VARIABLE FOR THE TARGET POPULATION SIZE
# SAVE AS: us_adult_pop_2020
# SET VALUE: 240000000
us_adult_pop_2020 <- 240000000
```

```{r}
# CALCULATE ADJUSTED WEIGHTS USING THE TARGET POPULATION
# FORMULA: (V200010b / sum(V200010b)) * us_adult_pop_2020
# SAVE AS: adjusted_weights
adjusted_weights <- (anes_data$V200010b / sum(anes_data$V200010b)) * us_adult_pop_2020
```

```{r}
# ADD THE ADJUSTED WEIGHTS TO THE DATASET
# SAVE AS: weight_adj (as new column in anes_data)
anes_data$weight_adj <- adjusted_weights
```

```{r}
# CREATE THE ANES SURVEY DESIGN OBJECT
# SAVE AS: anes_design
# USE: 
# - weights = weight_adj
# - strata = V200010d
# - ids = V200010c
# - nest = TRUE
anes_design <- anes_data %>%
  as_survey_design(
    weights = weight_adj,
    strata = V200010d,
    ids = V200010c,
    nest = TRUE
  )
```

```{r}
# EXAMINE THE SURVEY DESIGN OBJECT
# DISPLAY: anes_design
anes_design
```

### 3. Survey-Weighted Counts and Totals

Calculate basic counts and totals using survey weights.

```{r}
# CALCULATE THE ESTIMATED TOTAL NUMBER OF RESPONDENTS
# USE: survey_count() function
# DISPLAY: result with comma formatting
anes_design %>%
    survey_count() %>%
    knitr::kable(format.args = list(big.mark = ","))
```

When you display the estimated counts for the target population, we recommend rounding the estimates to make interpretation easier. Otherwise you have to explain why there are fractional people.

Additionally, remove missing values before performing the counting operation.

```{r}
# CALCULATE THE ESTIMATED COUNT BY EDUCATIONAL ATTAINMENT
# SAVE AS: education_counts
# DISPLAY: with comma formatting
education_counts <- anes_design %>%
    filter(!is.na(Education)) %>%
    survey_count(Education)

education_counts %>%
  knitr::kable(
    digits = 0,
    format.args = list(
        big.mark = ",",
        nsmall = 0
        )
    )
```



```{r}
# CALCULATE THE ESTIMATED COUNT BY EDUCATION AND GENDER
# USE: survey_count() function
# DISPLAY: with comma formatting
anes_design %>%
  filter(!is.na(Gender) & !is.na(Education)) %>%
  survey_count(Gender, Education, name = "N") %>%
  knitr::kable(
    digits = 0,
    format.args = list(
        big.mark = ",",
        nsmall = 0
        )
    )
```

```{r}
# REDO THE ABOVE ANALYSIS BUT PIVOT RESULTS WIDER
anes_design %>%
    filter(!is.na(Gender) & !is.na(Education)) %>%
    survey_count(Gender, Education, name = "N") %>%
    pivot_wider(
        names_from = Gender,
        values_from = c(N, N_se)
    ) %>%
    knitr::kable(
        digits = 0,
        format.args = list(
            big.mark = ",",
            nsmall = 0
        )
    )
```

### 4. Survey-Weighted Means and Proportions

Calculate means for continuous variables and proportions for categorical variables.

```{r}
# CALCULATE THE PROPORTION OF RESPONDENTS IN EACH EDUCATION LEVEL
# USE: survey_prop() function
# DISPLAY: with 3 decimal places
anes_design %>%
  filter(!is.na(Education)) %>%
  group_by(Education) %>%
  summarize(prop = survey_prop()) %>%
  knitr::kable(digits = 3)
```

```{r}
# CALCULATE THE MEAN AGE FOR THE ENTIRE POPULATION
# SAVE RESULT AS: mean_age
# DISPLAY: mean_age
mean_age <- anes_design %>%
  summarize(mean_age = survey_mean(Age, na.rm = TRUE))

mean_age
```

```{r}
# CALCULATE THE MEAN AGE BY GENDER
# SHOW CONFIDENCE INTERVAL
# DISPLAY: with 2 decimal places
anes_design %>%
    filter(!is.na(Gender)) %>%
    group_by(Gender) %>%
    summarize(
        mean_age = survey_mean(Age, vartype = "ci", na.rm = TRUE)
    ) %>%
    knitr::kable(digits = 2)
```

```{r}
# CALCULATE THE PROPORTION WHO VOTED IN 2020
# DISPLAY: results with standard errors
anes_design %>%
  filter(!is.na(VotedPres2020)) %>%
  group_by(VotedPres2020) %>%
  summarize(prop = survey_prop()) %>%
  knitr::kable(digits = 3)
```

### 5. Survey-Weighted Quantiles

```{r}
# CALCULATE THE MEDIAN AGE
# SAVE AS: median_age
# DISPLAY: median_age
median_age <- anes_design %>%
  summarize(
    median_age = survey_quantile(
        Age, 
        quantiles = 0.5, 
        na.rm = TRUE
        )
    )

median_age %>%
    knitr::kable()
```

```{r}
# CALCULATE THE 25th, 50th, AND 75th PERCENTILES OF AGE
# INCLUDE SEs 
# DISPLAY: results with proper labeling
anes_design %>%
    summarize(
        age_quartiles = survey_quantile(
            V201507x,
            quantiles = c(0.25, 0.5, 0.75),
            na.rm = TRUE
            )
        ) %>%
  knitr::kable(digits = 1)
```

```{r}
# CALCULATE AGE QUARTILES BY EDUCATION LEVEL
# DISPLAY: results with 1 decimal place
anes_design %>%
    filter(!is.na(Education)) %>%
    group_by(Education) %>%
    summarize(
        age_quartiles = survey_quantile(
            Age, 
            quantiles = c(0.25, 0.5, 0.75), 
            na.rm = TRUE
        )
    ) %>%
    knitr::kable(digits = 1)
```

### 6. Survey-Weighted Ratios

```{r}
# CALCULATE THE RATIO OF MEAN AGE BETWEEN COLLEGE GRADS AND NON-GRADS
# HINT: First create a binary education variable 
# - (college vs non-college)

# FIRST COMPARE [Education] vs. [V201510]
with(anes_design$variables, table(Education, V201510))
```

```{r}
# NOW UTILIZE [V201510] TO BUILD THE BINARY VARIABLE
# AND GET MEAN AGE BY COLLEGE OR PRE_COLLEGE
age_by_education <- anes_design %>%
    filter(!is.na(V201510)) %>%
    mutate(
    college_grad = ifelse(
        V201510 >= 4, 
        "College+", 
        "Less than College"
        )
    ) %>%
    group_by(college_grad) %>%
    summarize(mean_age = survey_mean(Age, na.rm = TRUE))

# VIEW THE DATAFRAME WITH THE MEAN AGES
age_by_education %>%
  knitr::kable(digits = 2)
```

```{r}
# CALCULATE RATIO
college_ratio <- age_by_education$mean_age[age_by_education$college_grad == "College+"] / 
                 age_by_education$mean_age[age_by_education$college_grad == "Less than College"]
print(paste("Ratio of college to non-college mean age:", round(college_ratio, 3)))
```

### 7. Comparison: Weighted vs Unweighted Analysis

Compare survey-weighted results with unweighted results to see the difference.

```{r}
# CALCULATE UNWEIGHTED MEAN AGE
# SAVE AS: unweighted_mean_age
# DISPLAY: result
unweighted_mean_age <- mean(anes_data$V201507x, na.rm = TRUE)
unweighted_mean_age
```

```{r}
# CALCULATE WEIGHTED MEAN AGE
# SAVE AS: weighted_mean_age  
# DISPLAY: result
weighted_mean_age <- anes_design %>%
  summarize(mean_age = survey_mean(V201507x, na.rm = TRUE))
weighted_mean_age
```

```{r}
# COMPARE THE DIFFERENCE BETWEEN WEIGHTED AND UNWEIGHTED ESTIMATES
# CALCULATE: difference between weighted and unweighted means
# DISPLAY: the difference and discuss what this suggests
age_difference <- weighted_mean_age$mean_age - unweighted_mean_age
print(paste("Difference (weighted - unweighted):", round(age_difference, 2), "years"))
print("This difference suggests that the survey design affects age estimates.")
```

```{r}
# CALCULATE UNWEIGHTED EDUCATION PROPORTIONS
# DISPLAY: results rounded to 3 decimal places
unweighted_education <- table(anes_data$Education, useNA = "no") %>%
    prop.table()
unweighted_education %>%
    knitr::kable(digits = 3)
```

```{r}
# CALCULATE WEIGHTED EDUCATION PROPORTIONS
# DISPLAY: results rounded to 3 decimal places
weighted_education <- anes_design %>%
  filter(!is.na(Education)) %>%
  group_by(Education) %>%
  summarize(weighted_prop = survey_prop())

print("Weighted Education Proportions:")
weighted_education %>%
  knitr::kable(digits = 3)

print("Comparison shows how survey weights adjust for population representation.")
```

```{r}
# BONUS - PRESENT WEIGHTED AND UNWEIGHTED RESULTS SIDE-BY-SIDE
# PUT THEM BOTH IN THE SAME TABLE

# Join results into a single table
comparison_table <- unweighted_education %>%
    as.data.frame() %>%
    rename(Education = Var1, unweighted_prop = Freq) %>%
    left_join(weighted_education, by = "Education") %>%
    select(Education, unweighted_prop, everything())

# Display the table
comparison_table %>%
    knitr::kable(
        digits = 2, 
        caption = "Weighted and Unweighted Mean Age by Education"
    )
```

## Conclusion

In this lab, you practiced computing survey-weighted descriptive statistics using the ANES 2020 dataset. You learned to calculate counts, means, proportions, quantiles, and ratios while properly accounting for the complex survey design through weights, strata, and clusters.

Key takeaways:
- Survey-weighted estimates provide population-representative results
- Unweighted analysis can lead to biased estimates that don't reflect the target population
- The survey design object is essential for proper variance estimation
- Different types of descriptive statistics require different survey functions

### Summary of Functions Used
- `survey_count()` and `survey_tally()` for weighted counts
- `survey_mean()` for weighted means 
- `survey_prop()` for weighted proportions
- `survey_quantile()` for weighted percentiles
- `survey_ratio()` for ratios between estimates
