---
title: "Survey Analysis with R - Inferential Statistics Lab SOLUTIONS"
author: "Instructor Solutions"
format: pdf
---

## Survey Inferential Statistics Lab
---

In this lab, you will practice conducting inferential statistical tests with survey data using the ANES 2020 dataset. You will learn to perform survey-adjusted hypothesis tests including t-tests, proportion tests, and chi-square tests while properly accounting for the complex survey design.

### 1. Load Libraries and Data

Load the required packages and set up the survey design object.

```{r}
# LOAD REQUIRED LIBRARIES
library(tidyverse)
library(survey)
library(srvyr)
```

```{r}
# IMPORT THE ANES 2020 DATA FROM THE DATA FOLDER
# SAVE AS: anes_data
anes_data <- read.csv('../data/anes_2020.csv')
```

```{r}
# CREATE TARGET POPULATION SIZE AND ADJUSTED WEIGHTS
# SET: us_adult_pop_2020 = 240000000
# CALCULATE: adjusted_weights using formula (V200010b / sum(V200010b)) * us_adult_pop_2020
# ADD: weight_adj column to anes_data
us_adult_pop_2020 <- 240000000
adjusted_weights <- (anes_data$V200010b / sum(anes_data$V200010b)) * us_adult_pop_2020
anes_data$weight_adj <- adjusted_weights
```

```{r}
# CREATE THE ANES SURVEY DESIGN OBJECT
# SAVE AS: anes_design
# USE: weights = weight_adj, strata = V200010d, ids = V200010c, nest = TRUE
anes_design <- anes_data %>%
  as_survey_design(
    weights = weight_adj,
    strata = V200010d,
    ids = V200010c,
    nest = TRUE
  )
```

### 2. One-Sample t-Tests

```{r}
# TEST WHETHER THE MEAN AGE DIFFERS FROM 45 YEARS
# USE: svyttest() with formula [Age] - 45 ~ 0
# SAVE AS: age_one_sample_test
# DISPLAY: test results
age_one_sample_test <- anes_design %>%
  filter(!is.na(Age)) %>%
  svyttest(formula = Age - 45 ~ 0, design = .)

age_one_sample_test
```

```{r}
# EXTRACT THE CONFIDENCE INTERVAL FOR THE AGE TEST
# USE: confint() function on the test result
# DISPLAY: confidence interval
confint(age_one_sample_test)
```

### 3. Two-Sample t-Tests

Compare means between two groups.

```{r}
# COMPARE MEAN AGE BETWEEN MALES AND FEMALES
# USE A HYPOTHESIS TEST
# USE: svyttest() with formula [Age] ~ [Gender]
# FILTER: remove missing values
# SAVE AS: age_gender_test
# DISPLAY: test results
age_gender_test <- anes_design %>%
  filter(!is.na(Age), !is.na(Gender)) %>%
  svyttest(formula = Age ~ Gender, design = .)

age_gender_test
```

```{r}
# CALCULATE GROUPED MEANS FOR AGE BY GENDER
# SHOW CONFIDENCE INTERVALS FOR THE GROUP MEANS
# GROUP BY: [Gender]
# DISPLAY: results with 2 decimal places
anes_design %>%
  filter(!is.na(Age), !is.na(Gender)) %>%
  group_by(Gender) %>%
  summarize(mean_age = survey_mean(Age, vartype = "ci")) %>%
  knitr::kable(digits = 2)
```

### 4. One-Sample Proportion Tests

```{r}
# EXAMINE THE PROPORTION WHO VOTED IN 2020
# GROUP BY: [VotedPres] (voting status)
# USE: survey_prop() to see current proportions
# FILTER: remove missing values
anes_design %>%
  filter(!is.na(VotedPres2020)) %>%
  group_by(VotedPres2020) %>%
  summarize(prop = survey_prop()) %>%
  knitr::kable(digits = 3)
```

```{r}
# TEST WHETHER THE VOTING RATE DIFFERS FROM 70%
# CREATE: voted variable (1 if [VotedPres] == 1, 0 otherwise)
# USE: svyttest() with formula (voted - 0.70) ~ 0
# DISPLAY: test results
voting_prop_test <- anes_design %>%
  filter(!is.na(VotedPres2020)) %>%
  mutate(voted = ifelse(VotedPres2020 == 'Yes', 1, 0)) %>%
  svyttest(formula = voted - 0.70 ~ 0, design = .)

voting_prop_test
```

```{r}
# EXTRACT CONFIDENCE INTERVAL FOR VOTING PROPORTION DIFFERENCE
# USE: confint() on the proportion test
# DISPLAY: confidence interval
confint(voting_prop_test)
```

### 5. Two-Sample Proportion Tests

```{r}
# COMPARE VOTING RATES BETWEEN COLLEGE AND NON-COLLEGE GRADUATES
# CREATE: college_grad and voted variables
# USE: svyttest() with formula voted ~ college_grad
# DISPLAY: test results
voting_education_test <- anes_design %>%
  filter(!is.na(VotedPres2020), !is.na(V201510)) %>%
  mutate(
    voted = ifelse(VotedPres2020 == 'Yes', 1, 0),
    college_grad = ifelse(V201510 >= 4, 1, 0)
  ) %>%
  svyttest(formula = voted ~ college_grad, design = .)

voting_education_test
```

```{r}
# CALCULATE VOTING PROPORTIONS BY EDUCATION LEVEL
# GROUP BY: education categories
# CALCULATE: proportion who voted in each group
# DISPLAY: results with confidence intervals
anes_design %>%
  filter(!is.na(VotedPres2020), !is.na(V201510)) %>%
  mutate(
    voted = ifelse(VotedPres2020 == 'Yes', 1, 0),
    college_grad = ifelse(V201510 >= 4, 1, 0)
  ) %>%
  group_by(college_grad) %>%
  summarize(voting_rate = survey_mean(voted, vartype = "ci")) %>%
  knitr::kable(digits = 3)
```

### 6. Chi-Square Tests of Independence

```{r}
# TEST INDEPENDENCE BETWEEN EDUCATION LEVEL AND VOTING BEHAVIOR
# USE: svychisq() with formula ~[Education] + [VotedPres2020]
# FILTER: remove missing values from both variables
# SAVE AS: education_voting_test
# DISPLAY: test results
education_voting_test <- anes_design %>%
  filter(!is.na(Education), !is.na(VotedPres2020)) %>%
  svychisq(formula = ~ Education + VotedPres2020, design = .)

education_voting_test
```

```{r}
# CREATE A CROSSTAB OF EDUCATION AND VOTING
# USE: survey functions to create weighted crosstab
# GROUP BY: both variables
# CALCULATE: survey_prop() for each combination
# DISPLAY: results in table format
anes_design %>%
  filter(!is.na(Education), !is.na(VotedPres2020)) %>%
  group_by(Education, VotedPres2020) %>%
  summarize(prop = survey_prop()) %>%
  knitr::kable(digits = 3)
```

### 7. Goodness-of-Fit Tests

```{r}
# EXAMINE THE ESTIMATED EDUCATION DISTRIBUTION FIRST
# GROUP BY: [Education]
# USE: survey_prop() to see actual distribution
# DISPLAY: estimated regional proportions
anes_design %>%
  filter(!is.na(Education)) %>%
  group_by(Education) %>%
  summarize(prop = survey_prop()) %>%
  knitr::kable(digits = 3)
```

```{r}
# PERFORM GOODNESS-OF-FIT TEST FOR EDUCATION DISTRIBUTION
# COMPARE: against expected proportions 
# Bachelor's - 0.3
# Graduate - 0.10
# High School - 0.2
# Less than HS - 0.1
# Post HS - 0.3
# SAVE AS: region_gof_test
# DISPLAY: test results
region_gof_test <- anes_design %>%
  filter(!is.na(Education)) %>%
  svygofchisq(formula = ~ Education, 
              design = .,
              p = c(.3,.1,.2,.1,.3))  # Expected proportions

region_gof_test
```

## Conclusion

In this lab, you practiced applying inferential statistical methods to complex survey data using the ANES 2020 dataset. You learned how to conduct survey-adjusted t-tests, proportion tests, and chi-square tests, and how to interpret the results in the context of survey sampling. These skills are essential for drawing valid conclusions from survey data, accounting for weights, stratification, and clustering. Continue to explore these methods with other variables and datasets to deepen your understanding of survey inference in R.